---
output:
  pdf_document:
    keep_tex: true
    fig_caption: true
    toc: true
    toc_depth: 3 
    number_sections: true
    citation_package: natbib
    latex_engine: pdflatex
    template: ./../latex-templates/report.tex
  html_document:
    df_print: paged
    
title: "The Influence of Biological Sex on Human Body Part Ratios"
subtitle: "How these Ratios Compare Between the Sexes"
author: 
- name: "Kevin A. Black - kevin.black@wsu.edu"
  affiliation: "Washington State University Vancouver"
keywords: |
    multiple comparisons to control; multivariate chi-square distribution; nonlinear growth curves; Richard's curve; simulated critical points
abstract: |
  In this article we compare the \emph{empirical characteristic function} \citep{Tukey:1977, Becker:1988} to a \emph{moment-generating-functional form} to compute the proportion of hypotheses $m$ that are rejected under the null hypothesis. \vspace{0.25in}

  \noindent Here is a second paragraph of the abstract (if necessary), and with the pipe notation it doesn't break. Notice it still needs to be indented. \vspace{0.25in}

  \noindent Generally, we write this abstract last.  Often it is called the executive summary. It should succinctly summarize the entire document.  You can include references such as this one to the Appendices section \ref{sec:appendix} if necessary.
sectionnumberdepth: 3
titleEndnotes: "ENDNOTES"
titleReferences: "REFERENCES"
columnsReferences: 2
titleTOC:  "TABLE OF CONTENTS"
bibliography: ./../biblio/master.bib
bibliostyle: ./../biblio/ormsv080.bst
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section{Introduction}
\label{sec:intro}

As a customary portion of the STAT 419 course at Washington State University, the students involved were required to take part in a data collection/manipulation project involving the measurements of various body parts. Following the data collection, professor Monte Shaffer compiled all contributions made by the students into a single, pipe-delimited text file whose explicit content is to remain confidential and only used for the sake of addressing a variety of research questions.

\vspace{0.25cm}

After receiving the data set ($n=428$) and cleaning up particular observations using a range of methodologies based on what I had in mind for how this data will be used, I determined that I would focus on what kind of influence biological sex has on certain human body part ratios. To supplement my inquiry of this data, I centered my research questions around particular measurements of the human body, such as height, arm span, foot length, distance from elbow to armpit, head height, just to name a few. Research questions pertaining to this study and supportive R code can be found below.

\section{How does being male or female influence the ratios between certain body parts?}
\label{sec:rq}

\subsection{How does 'height' compare to 'arm span' between males and females?}
\label{sec:rq2}

\subsection{How does 'foot length' compare to the length of 'elbow to armpit' between males and females?}
\label{sec:rq3}

\subsection{How many average 'head height' lengths are males and females relative to their respective, average 'height'?}
\label{sec:rq4}

\section{Data Description}
\label{sec:data}

As mentioned in the introduction, the data set utilized throughout this research paper was supplied through the combined efforts of professor Monte Shaffer and the students of the STAT 419 class at Washington State University. Each student was tasked with recording distinct observations of 37 attributes from 10 different people, ideally with an even mix of males and females. For each observation (person), there are measurements of body parts, data collector/respondent identifiers ran through a MD5 hash function, and general information about each respondent. Body part measurements were recorded using body measuring tape in either inches or centimeters, but for the sake of consistency throughout the research paper, all centimeter values were converted to inches and will be treated as inches from here on.

\vspace{0.25cm}

These observations were recorded in early September 2020 and compiled by the instructor for our use in late October 2020. Given that these observations were recorded amid the COVID-19 pandemic, each student was required to make a simple, yet descriptive handout that would detail how one would go about recording their own body measurements and the necessary values to take note of. This would be an ideal situation of how observations were recorded and sent electronically by each respondent, but observations could also be taken in person given that the surveyor and respondent were comfortable being in close proximity of each other. An example of a two-page handout created by Kevin Black, as well as further information regarding this data set, can be found in Appendix \ref{sec:appendix-data-handout} and Appendix \ref{sec:appendix-dataset-ex}, respectively.

\vspace{0.25cm}

On the surface, the purpose for writing this research paper and collecting the necessary data can simply be attributed to project requirements for a university course. However, the deeper reasoning behind why this research paper was composed the first place was to give students a more thorough understanding of the data analytics process. More specifically, how to not just work with the data, but how to understand the data and derive effective questions, how to test for patterns and conclusions in a statistical, analytic environment, and how to exercise data provenance practices. This process will be very similar between all data focused projects in a data analyst's career, so this is a good starting point for garnering crucial experience. 

\section{Key Findings}
\label{sec:findings}

\section{Conclusion}
\label{sec:conclusion}

\newpage

\input{tables/example-correlation-table} 

\newpage

\section{APPENDICES}
\label{sec:appendix}

\subsection{Data Provenance}
\label{sec:appendix-data-provenance}

\subsubsection{Utilization of Data Provenance}
\label{sec:appendix-provenance-explained}

\newpage

\subsubsection{Data Collection Handout}
\label{sec:appendix-data-handout}

\begin{figure}[!ht]
	\hrule
	\caption{ \textbf{Handout Page 1} }
	\begin{center}
	    \scalebox{1.00}{	\includegraphics[trim = 0 0 0 0,clip,width=0.85\textwidth]{pdfs/handout1.pdf} }
	\end{center}
	\label{fig:handout-1}
	\hrule
\end{figure}

\newpage

\begin{figure}[!ht]
	\hrule
	\caption{ \textbf{Handout Page 2} }
	\begin{center}
	    \scalebox{1.00}{	\includegraphics[trim = 0 0 0 0,clip,width=0.85\textwidth]{pdfs/handout2.pdf} }
	\end{center}
	\label{fig:handout-2}
	\hrule
\end{figure}

\newpage

\subsection{Data Set Explained}
\label{sec:appendix-dataset-ex}
In addition to how the data set was described in Section \ref{sec:data}, explanations regarding each attribute can be found below. After the collaborative effort of all students having their data compiled, the data set ended up having 428 total observations. However, the data set was filled with an enormous amount of NA values in the body measurements, potentially due to the time constraints of some students or lack of attempt to fill out all fields. As a result, running the function `complete.cases()` on the data set during the data cleaning process returned a data frame containing only 262 observations, about 61.21% of the original data set size. `complete.cases()` was used over `na.omit()` because the latter function would omit all rows containing NA values based on all columns, including the rows where the non-body measurement attributes had NA values, whereas the former function would omit NA values only for the specified body measurement columns.

\begin{figure}[!ht]
	\caption{ \textbf{Description of Each Field} }
	\begin{center}
	    \scalebox{1.00}{	\includegraphics[trim = 0 350 100 35,clip,width=0.85\textwidth]{pdfs/datasets_explained.pdf} }
	\end{center}
	\label{fig:datasets_explained}
\end{figure}

\newpage

\subsection{R Code Used for Research}
\label{sec:appendix-setup}

Below is the necessary functions and libraries required to run the code referenced in this document.

```{r,fig.show="hide",message=FALSE}
# Import necessary libraries
library(stats) # For cor()
library(devtools) # For source_url()

# Source cleaning function
path.hub = "https://raw.githubusercontent.com/KevnBlack/WSU_STATS419_FALL2020/"
source_url(paste0(path.hub,"master/functions/functions-project-measure.R"))

# Import data
path.to.secret = "D:/School/Fall 2020/STAT 419/datasets/"
measure.df = read.file(path.to.secret)

# Isolate male and female data
measure.df.m = measure.df[measure.df$gender == "male",]
measure.df.f = measure.df[measure.df$gender == "female",]
cor.m = cor(measure.df.m$height.NA, measure.df.m$arm.span.NA)
cor.f = cor(measure.df.f$height.NA, measure.df.f$arm.span.NA)

# Correlation Graphs
par(mfrow = c(1,2))
plot(measure.df.m$height.NA, measure.df.m$arm.span.NA,
     main = paste("Correlation =", round(cor.m,5)),
     xlab = "Male Height (in)", ylab = "Male Arm Span (in)", col = "blue")

plot(measure.df.f$height.NA, measure.df.f$arm.span.NA,
     main = paste("Correlation =", round(cor.f,5)),
     xlab = "Female Height (in)", ylab = "Female Arm Span (in)", col = "red")

```






